---
title: "MALTEM maps"
author: "Joe Brew and Bea Galatas"
date: "March 23, 2016"
output: pdf_document
---  


# Preparation   

Prepare libraries  
```{r, results = 'hide'}
library(raster)
library(maptools)
library(sp)
library(mgcv)
library(plyr)
library(readr)
library(rgdal)
library(ggmap)
library(ggplot2)
library(grid)
library(dplyr)
library(knitr)
library(ggthemes)
library(RColorBrewer)
library(akima)
```

Get spatial data

```{r}

#### MOZAMBIQUE
# Get a shapefile for Mozambique
moz <- raster::getData("GADM", country = "MOZ", level = 3)

#### MAGUDE
# Read in shapefile of Magude
magude <- readOGR("Adm3", "Magude_adm3")
# Ensure that only the relevant data
# remains in the spatial object:
magude <- magude[magude@data$NAME_2 == "Magude",]
# Fortify magude (format for ggplot2)
magude_fortified <- 
  fortify(magude, region = 'NAME_3')
# Get lng and lat in magude@data
magude@data$lng <- coordinates(magude)[,1]
magude@data$lat <- coordinates(magude)[,2]

#### CMAGVEC : CENSUS VECTOR CONTROL DATA
#Importing Census vector control data
cmagvec <-read_csv("household_vector_gps.csv")
cmagvec <- na.omit(data.frame(cmagvec))
cmagvec$irs[cmagvec$irs==2] <- NA

# Make lat/lng columns in cmagvec
cmagvec$latitude <- cmagvec$lat
cmagvec$longitude <- cmagvec$lng

# Create a boolean for irs, atl1net, univnet, allsleptnet
# atl1slept
booleans <- c('irs', 
              'atl1net',
              'univnet',
              'allsleptnet',
              'atl1slept')
for(j in booleans){
  print(j)
  cmagvec[,paste0(j, '_boolean')] <- NA
  cmagvec[,paste0(j, '_boolean')] <- as.logical(cmagvec[,j])
}

# Create and id row
cmagvec$id <- 1:nrow(cmagvec)

# Create a spatial version of cmagvec
coordinates(cmagvec) <- ~longitude+latitude

# Get satellite map
if('tempmap.RData' %in% dir('/tmp')){
  load('/tmp/tempmap.RData')
} else {
  sat_map <- get_map(location = as.numeric(apply(bbox(cmagvec), 1, mean)),
                  maptype = 'satellite', 
                  zoom = 9)
  save('sat_map', file = '/tmp/tempmap.RData')
}
```

# Create raster file

```{r}
#### Basic method:
# first perform spatial interpolation, 
# then rasterize

# Create a temporary object with the values of 
# irs (averaging for shared points)
temp <- cmagvec@data %>%
  group_by(lng, lat) %>%
  summarise(irs = mean(irs, na.rm = TRUE)) %>%
  mutate(irs = ifelse(is.na(irs), 1, irs))

steps <- 500
isu <- with(temp, 
            akima::interp(lng, lat, irs, 
                   xo=seq(min(lng), max(lng), length = steps),
                   yo=seq(min(lat), max(lat), length = steps),
                   # extrap = TRUE,
                   linear = TRUE))
r <- raster(isu)
plot(magude)
plot(r, add = TRUE)
plot(magude, add = T)
```

The above is less than satisfactory - not smoothed enough. So, let's create our own dataframe with our custom interpolation method

```{r}
# Create a gridded dataframe with values 
# for the entire range (bbox) of magude
df_grid <- expand.grid(lng = seq(bbox(magude)[1,1],
                                 bbox(magude)[1,2],
                                 by = 0.01),
                       lat = seq(bbox(magude)[2,1],
                                 bbox(magude)[2,2],
                                 by = 0.01),
                       irs = NA,
                       irs_score = NA,
                       color = NA)
df_grid$latitude <- df_grid$lat
df_grid$longitude <- df_grid$lng
coordinates(df_grid) <- ~longitude+latitude

# Create a 1001 length color vector
color_vector <-
  colorRampPalette(c('red', 'yellow', 'green'))(1001)

# Go through each row of df_grid, getting the 
# weighted mean irs score for that point
# and putting a color into df_grid
if('df_grid.RData' %in% dir()){
  load('df_grid.RData')
} else {
  for (i in 1:nrow(df_grid)){
  # Get distance from every point in cmagvec
  distances <- spDistsN1(pts = cmagvec,
                        pt = df_grid[i,],
                        longlat = TRUE)
  # Define which are acceptably close
  close_enough <- which(distances <= 50)
  # Get an IRS score
  irs <- stats::weighted.mean(x = cmagvec$irs[close_enough],
                       w = (1 / distances[close_enough]) ^2,
                       na.rm = TRUE)
  # Assign irs to the dataframe
  df_grid$irs[i] <- irs
  # Get the IRS score on a 0-1000 scale
  # and add 1 so that it's on a 1-1001 scale (for coloring)
  irs_score <- round(irs * 1000) + 1
  df_grid$irs_score[i] <- irs_score
  # If there are no nearby obs, just make white
  # otherwise, give it a color from the palette
  if(!is.na(irs_score)){
   color <- color_vector[round(irs_score )] 
  } else {
    color <- 'white'
  }
  df_grid$color[i] <- color
}
save('df_grid', file = 'df_grid.RData')

}

plot(df_grid, col = df_grid$color, 
     pch = 16,
     cex = 0.4)
plot(magude, add = T)
```

Much better!  Now that we like the "smoothness" of our interpolation, let's make our object (currently a "SpatialPointsDataFrame") into a raster object.

```{r}
# Convert df_grid to raster
temp <- df_grid@data %>% arrange(lng, lat)
r <- rasterFromXYZ(temp[, c('lng', 'lat', 'irs')])
plot(r)
```

If we want, we can make our grid smaller, so that it only contains points within the Magude district:

```{r}
proj4string(df_grid) <- proj4string(magude)
x <- over(df_grid, polygons(magude))
df_grid_small <- df_grid[!is.na(x),]
plot(df_grid_small,
     col = adjustcolor('black', alpha.f = 0.2))
```

That way, instead of having surfaces like this:
```{r}
plot(r)
```

We can instead get surfaces like this:

```{r}
temp <- df_grid_small@data %>% arrange(lng, lat)
r <- rasterFromXYZ(temp[, c('lng', 'lat', 'irs')])
plot(r)
```

Raster objects are cool, because we can plot them in different ways:

```{r}
plot(magude)
contour(r, add = TRUE,
        col = adjustcolor('black', alpha.f = 0.3))

plot(magude)
plot(r, add = TRUE)
plot(magude, add = TRUE)

plot(magude)
image(r, add = TRUE)
```

We can also plot our gridded creation using the ggplot framework.

```{r}
ggplot(magude_fortified,
       aes(x = long, y = lat)) +
  geom_polygon(aes(group = id), fill = 'black') +
  geom_point(data = temp,
             aes(x = lng, y = lat, col = irs)) +
    geom_polygon(aes(group = id), fill = NA, color = 'black', alpha = 0.5) +
  coord_map() +
    xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('IRS coverage') +
  theme(panel.background = element_rect(fill = 'white', 
                                        colour = 'white'),
        panel.grid.major = element_line(colour = "white"))
```


# Other maps  

Using ggplot, we can make pretty point maps as well.

```{r}
# Define a two-color palette
cols2 <- c('red', 'lightblue')
cols2 <- adjustcolor(cols2, alpha.f = 0.4)

# Make map
ggplot(magude_fortified,
       aes(x = long, y = lat)) +
  geom_polygon(aes(group = id), fill = 'black') +
  geom_polygon(aes(group = id), fill = NA, color = 'darkgrey', alpha = 0.2) +
  geom_point(data = cmagvec@data,
             aes(x = lng, y = lat, col = irs_boolean), size = 0.35) +
  coord_map() +
  scale_colour_manual(values = cols2, 
                      name = 'IRS') +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('IRS coverage') +
  # element_blank() +
  # theme(panel.background = element_rect(colour = "black")) +
  theme(panel.background = element_rect(fill = 'white', 
                                        colour = 'white'),
        panel.grid.major = element_line(colour = "white"))
```

Finally, we can use the ggmaps package to visualize our data over satellite imagery

```{r}
ggmap(sat_map) +
  geom_polygon(data = magude_fortified,
       aes(x = long, y = lat, group = id), 
       fill = NA,
       color = 'darkgrey') +
  coord_map() +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('IRS coverage') +
  theme(panel.background = element_rect(fill = 'white', 
                                        colour = 'white'),
        panel.grid.major = element_line(colour = "white")) +
  geom_tile(data = temp,
            aes(x = lng, y = lat, fill = irs), alpha = 0.7) +
  scale_fill_gradient(low = "red", high = "white")
```

# Polygonal manipulation  

Rasters are continuous surfaces.  Polygons are "zones".  Magude has 5 (Panjane, Magude, Mahele, Mapulganguene, Motaze).  We can examine average IRS coverage by zone.

```{r}
# Ensure that cmagvec shares projection information with magude
proj4string(cmagvec) <- proj4string(magude)

# Get the zone of each unit in the census
x <- over(cmagvec, polygons(magude))
cmagvec$zone <- magude@data$NAME_3[as.numeric(x)]

# Get average IRS coverage by zone
by_zone <- cmagvec@data %>%
  group_by(zone) %>%
  summarise(irs = 100 * mean(irs, na.rm = TRUE)) %>%
  filter(!is.na(zone))

# Join the zone data to the data component
# of the magude shapefile
magude@data <- 
  left_join(x = magude@data,
            y = by_zone,
            by = c('NAME_3' = 'zone'))

# Bring that data into the "fortified" version of magude
magude_fortified <- 
  left_join(x = magude_fortified,
            y = by_zone,
            by = c('id' = 'zone'))

# Use the ggplot package to create a choropleth map
ggmap(sat_map) +
  geom_polygon(data = magude_fortified,
       aes(x = long, y = lat, group = id, 
           fill = irs),
       color = 'darkgrey') +
  coord_map() +
  xlab('Longitude') +
  ylab('Latitude') +
  ggtitle('IRS coverage (polygonal)') +
  theme(panel.background = element_rect(fill = 'white', 
                                        colour = 'white'),
        panel.grid.major = element_line(colour = "white")) +
  scale_fill_gradient(low = "red", high = "white")

```
